---
title: CFA - Examples
bibliography: references.bib
---

```{r}
#| label: setup

# Load packages
library(tidyverse)
library(lavaan)
library(dynamic)
```

## Rosenberg Self-Esteem Scale

The Rosenberg self-esteem scale (Rosenberg, 1989) is a unidimensional scale to assess global self-esteem. It is one of the most commonly used scales in psychology and could therefore serve as a nice illustration of how to assess whether the data fits a single factor model.

It should be noted, however, that apparently there is some controversy whether the scale is actually unifactorial [@donnellan2016].

### The Data

```{r}
#| label: data

# Load data
items <- read_csv("RSECSV.csv")

# Data preparation
# Replace all -9999 values with NA
items <- mutate(items, across(everything(), ~ na_if(.x, -9999)))
```

The data was taken from @donnellan2016, who conducted a study to assess the factorial structure of the RSE. The data consists of `nrow(items)` participants, who were college students at a public university in the Southwestern United States.

### The Model

```{r}
#| label: model
model <- "
  self_esteem =~ R1 + R2 + R3 + R4 + R5 + R6 + R7 + R8 + R9 + R10
"

fit <- cfa(model = model, data = items, estimator = "MLR")
fitmeasures(fit, 
  fit.measures = c("pvalue", "pvalue.scaled", "srmr", "rmsea", "cfi"))
#fit_cutoffs <- cfaOne(fit)
```

## Marsh's Self-Description Questionnaire

```{r}
#| label: marsh-data

# Load data
marsh <- read_csv("marsh.csv")
```

Data for this example is from 15,661 students in 10th grade who participated in the National Education Longitudinal Study of 1988. Items from this scale include:

Parents:

-   My parents treat me fairly

-   I do not like my parents very much

-   I get along well with my parents

-   My parents are usually unhappy or disappointed with what I do

-   My parents understand me

English

-   I learn things quickly in English classes

-   English is one of my best subjects

-   I get good marks in English

-   I'm hopeless in English classes

Mathematics

-   Mathematics is one of my best subjects

-   I have always done well in mathematics

-   I get good marks in mathematics

-   I do badly in tests of mathematics

```{r}
#| label: marsh-model
marsh_model <- "
  english =~ Eng1 + Eng2 + Eng3 + Eng4
  math    =~ Math1 + Math2 + Math3 + Math4
  parent  =~ Par1 + Par2 + Par3 + Par4 + Par5
"

marsh_fit <- cfa(model = marsh_model, data = marsh, estimator = "MLR")
fitmeasures(marsh_fit, 
  fit.measures = c("pvalue", "pvalue.scaled", "srmr", "rmsea", "cfi")
)
#marsh_fit_cutoffs <- cfaHB(marsh_fit)
```
