---
title: "Comparing string similarity"
toc: true
---

```{r}
#| label: setup
#| message: false
library(tidyverse)
library(stringdist)
```

String similarity metrics reduce the comparison between two pieces of text to a single number, making it possible to rank, filter, or aggregate comparisons at scale. Common use cases include comparing survey responses across time points, detecting duplicate records, and evaluating translations. This page focuses on the translation use case — we have a reference text and want to measure how closely several translations match it — but the methods apply wherever you need to compare strings.

This page covers three general-purpose surface-level metrics from the `stringdist` package — Levenshtein, Jaccard, and cosine — BLEU, a metric designed specifically for translation evaluation, and semantic similarity using embeddings, which captures meaning rather than surface overlap. Each section shows how the metric works on simple examples before applying it to translations.

## Distance vs. similarity

Most string comparison algorithms compute a *distance* — how different two strings are. Raw distances are hard to interpret because they scale with string length: one changed character in a 4-character string is a much bigger deal than one changed character in a 100-character string.

*Similarity* normalizes distance into a 0–1 scale where 1 means identical and 0 means completely different. The `stringdist` package provides both `stringdist()` for raw distances and `stringsim()` for normalized similarities. This page uses `stringsim()` throughout.

## Example data

We use a single source sentence and four translations that vary in how closely they match. Before computing any similarity, it is usually worth cleaning the strings — lowercasing, stripping punctuation — so that differences in capitalization or punctuation do not distort the scores.

```{r}
#| label: data
clean <- function(x) x |> str_to_lower() |> str_remove_all("[[:punct:]]") |> str_squish()

source <- clean("The cat sat on the mat near the window.")

translations <- tribble(
  ~label,          ~text,
  "Near-identical","The cat sat on the mat near the window!",
  "Paraphrase",    "The cat was resting on the mat beside the window.",
  "Different",     "A dog lay on the floor next to the door.",
  "Unrelated",     "The economy grew faster than expected last quarter."
) |>
  mutate(text = clean(text))
```

## Levenshtein similarity

The Levenshtein distance asks: what is the smallest number of single-character edits — insertions, deletions, or substitutions — needed to turn one string into the other? `stringsim()` then normalizes this by dividing by the length of the longer string, so the result is always between 0 and 1.

```{r}
#| label: levenshtein-intro
stringsim("cat", "car", method = "lv")   # one substitution (t → r); longer string is 3 chars; sim = 1 - 1/3
stringsim("cat", "cart", method = "lv")  # one insertion (add t);    longer string is 4 chars; sim = 1 - 1/4
stringsim("cat", "dog", method = "lv")   # three substitutions;      longer string is 3 chars; sim = 1 - 3/3
stringsim("cat", "cat", method = "lv")   # identical; distance = 0;  sim = 1
```

The insertion case (`"cat"` vs `"cart"`) scores higher than the substitution case (`"cat"` vs `"car"`) even though both require exactly one edit. The reason is normalization: the longer string `"cart"` has 4 characters, so the penalty is 1/4. With `"car"`, the longest string has only 3 characters, so the same edit costs 1/3. The same number of edits looks smaller when the strings are longer.

Applied to the translations:

```{r}
#| label: levenshtein-translations
translations |>
  mutate(similarity = stringsim(text, source, method = "lv")) |>
  select(label, similarity)
```

## Jaccard similarity

Levenshtein cares about the exact sequence of characters. Jaccard takes a different approach: it ignores order entirely and just asks what words the two strings have in common.

Each string is treated as a bag of words. Jaccard similarity is the number of words that appear in both strings, divided by the total number of distinct words across both strings. If two strings share all their words, the score is 1. If they share none, it is 0.

```{r}
#| label: jaccard-intro
stringsim("the cat sat", "the cat ran", method = "jaccard")  # "the" and "cat" shared; "sat" and "ran" not
stringsim("the cat sat", "the cat sat", method = "jaccard")  # all words shared
stringsim("the cat sat", "a dog ran",   method = "jaccard")  # no words shared
```

Because word order does not matter, `"cat sat mat"` and `"mat cat sat"` are treated as identical. This makes Jaccard useful when a translation reorders or restructures a sentence while keeping the same vocabulary. It also means Jaccard is blind to repetition: a word that appears three times counts the same as one that appears once.

Applied to the translations:

```{r}
#| label: jaccard-translations
translations |>
  mutate(similarity = stringsim(text, source, method = "jaccard")) |>
  select(label, similarity)
```

## Cosine similarity

Cosine similarity counts how often each short character sequence appears in each string, then asks whether the two strings have similar distributions of those sequences. These sequences are called *character n-grams*: an n-gram is a sequence of n consecutive units, and here the units are characters. If the same character patterns show up with similar frequency in both strings, the score is high. If the patterns are very different, the score is low.

```{r}
#| label: cosine-intro
stringsim("hello world", "hello there", method = "cosine")   # "hel", "ell", "llo" etc. shared
stringsim("hello world", "goodbye world", method = "cosine") # "wor", "orl", "rld" shared
stringsim("hello world", "hello world", method = "cosine")   # identical
stringsim("hello world", "xyz abc", method = "cosine")       # no character patterns in common
```

The main advantage of cosine over Levenshtein is that it is not thrown off by length differences. A long paraphrase and a short original can still score high if they share the same character-level patterns, whereas Levenshtein would penalize the extra characters.

Applied to the translations:

```{r}
#| label: cosine-translations
translations |>
  mutate(similarity = stringsim(text, source, method = "cosine")) |>
  select(label, similarity)
```

## Comparing metrics

The three metrics often agree but can diverge:

- Levenshtein is sensitive to exact character-level edits, including small differences in spelling or word endings.
- Jaccard cares only about which words appear, not how many times or in what order.
- Cosine looks at the distribution of character patterns and is least sensitive to length differences.

Computing all three gives a fuller picture:

```{r}
#| label: all-metrics
translations |>
  mutate(
    levenshtein = stringsim(text, source, method = "lv"),
    jaccard     = stringsim(text, source, method = "jaccard"),
    cosine      = stringsim(text, source, method = "cosine")
  ) |>
  select(label, levenshtein, jaccard, cosine)
```

## BLEU score

The three metrics above are general-purpose string comparison tools. BLEU (Bilingual Evaluation Understudy) was designed specifically for evaluating translations and remains the standard metric in translation research. Its core question is different: rather than asking how similar two strings are character-by-character or word-by-word, it asks how much of the candidate translation's phrasing actually appears in the reference.

### The core idea: n-gram precision

The starting point is word-level precision: what fraction of the words in the candidate also appear in the reference?

```{r}
#| label: bleu-precision-intuition
ref  <- "the cat sat on the mat"
cand <- "the cat sat on the floor"
```

The candidate has 6 words. Five of them — "the", "cat", "sat", "on", "the" — appear in the reference. Only "floor" does not. Naïve precision would be 5/6 ≈ 0.83.

But precision alone misses something important: it only checks whether each word appears somewhere in the reference, not whether adjacent words appear *together*. A translation that gets every individual word right but scrambles them completely would score the same as one that is perfectly fluent. BLEU addresses this by checking not just individual words but consecutive sequences of words. These are also called n-grams — the same term used in the cosine section above, but there applied to characters rather than words. An n-gram is just a sequence of n consecutive units; what those units are (characters, words, something else) depends on the method.

It computes precision separately for unigrams (single words), bigrams (pairs), trigrams (triples), and 4-grams, then combines them. Getting a 4-gram right is a much stronger signal than getting a single word right: the candidate would have had to produce four words in exactly the right order. High n-gram precision at longer lengths means the translation is fluent, not just using the right vocabulary.

### Clipped precision

Naïve precision has a flaw. Consider a candidate that just repeats the most common word in the reference:

```{r}
#| label: bleu-clipping-problem
ref_words  <- str_split("the cat sat on the mat", " ")[[1]]
cand_words <- str_split("the the the the the the", " ")[[1]]

mean(cand_words %in% ref_words)  # every "the" matches; precision = 1.0
```

Every word in the candidate appears in the reference, so naïve precision is 1.0, even though the candidate is nonsense.

BLEU fixes this with *clipped precision*: each reference word can only be credited once. If "the" appears twice in the reference, it can match at most two "the"s in the candidate, no matter how many times the candidate repeats it.

```{r}
#| label: bleu-clipping-demo
ref_counts  <- table(ref_words)   # "the" appears 2 times in reference
cand_counts <- table(cand_words)  # "the" appears 6 times in candidate

# Clipped: each word can only be matched up to its count in the reference
clipped <- sum(map_dbl(names(cand_counts), \(w) {
  min(cand_counts[w], ref_counts[w] %||% 0)
}))

clipped / length(cand_words)  # 2 / 6 ≈ 0.33; much more honest
```

### Brevity penalty

Precision has a second weakness: shorter candidates tend to score higher because there are fewer n-grams to get wrong. An extreme case is a one-word candidate: if that word appears in the reference, unigram precision is 1.0.

BLEU applies a brevity penalty when the candidate is shorter than the reference. The penalty grows as the length gap grows. If the candidate is at least as long as the reference, no penalty applies.

```{r}
#| label: bleu-bp
brevity_penalty <- function(cand_tokens, ref_tokens) {
  if (length(cand_tokens) >= length(ref_tokens)) {
    1
  } else {
    exp(1 - length(ref_tokens) / length(cand_tokens))
  }
}
```

### Implementation

There is no widely used R package for BLEU that fits naturally into a tidyverse workflow, so we implement it directly. The three components — n-gram extraction, clipped precision, and the brevity penalty — map cleanly to three small functions.

```{r}
#| label: bleu-implementation
get_ngrams <- function(tokens, n) {
  if (length(tokens) < n) return(character(0))
  map_chr(seq_len(length(tokens) - n + 1), \(i) paste(tokens[i:(i + n - 1)], collapse = " "))
}

clipped_precision <- function(cand_tokens, ref_tokens, n) {
  cand_ngrams <- get_ngrams(cand_tokens, n)
  ref_ngrams  <- get_ngrams(ref_tokens, n)
  if (length(cand_ngrams) == 0) return(0)
  sum(map_dbl(unique(cand_ngrams), \(ng) {
    min(sum(cand_ngrams == ng), sum(ref_ngrams == ng))
  })) / length(cand_ngrams)
}

bleu <- function(candidate, reference, max_n = 4) {
  cand_tokens <- str_split(candidate, "\\s+")[[1]]
  ref_tokens  <- str_split(reference, "\\s+")[[1]]
  precisions  <- map_dbl(seq_len(max_n), \(n) clipped_precision(cand_tokens, ref_tokens, n))
  if (any(precisions == 0)) return(0)
  bp <- brevity_penalty(cand_tokens, ref_tokens)
  bp * exp(mean(log(precisions)))
}
```

The final score is the geometric mean of the four clipped precisions, multiplied by the brevity penalty. The geometric mean is used rather than the arithmetic mean because it returns 0 the moment any precision is 0 — a translation that completely fails on 4-grams should not score well just because its individual words are right.

### Applied to the translations

```{r}
#| label: bleu-translations
translations |>
  mutate(bleu = map_dbl(text, bleu, reference = source)) |>
  select(label, bleu)
```

A few things to notice in the output. The near-identical translation scores very high because its word sequences almost perfectly match the reference. The paraphrase scores lower even though it conveys a similar meaning — BLEU does not know that "resting" is similar to "sat" or that "beside" is similar to "near". The unrelated translation scores 0 because it shares no 4-grams with the reference, and a single zero precision collapses the entire score to 0.

### Limitations

BLEU measures surface overlap, not meaning. Two sentences that say the same thing in different words — "the cat sat on the mat" and "the feline rested upon the rug" — score near zero against each other, because they share almost no n-grams. This is the fundamental limitation of all the metrics on this page. For use cases where meaning matters more than exact wording, embedding-based metrics like BERTScore are more appropriate; they compare translations in a semantic space where "cat" and "feline" are close together.

BLEU also has a known sensitivity to short texts. On sentence-level comparisons, scores are noisier and harder to interpret than on full documents, because there are fewer n-grams to aggregate over. When comparing individual sentences, the `stringsim()` metrics are often more stable and easier to reason about; BLEU becomes more reliable when aggregated over many sentences or paragraphs.

## Semantic similarity

All the metrics above measure surface overlap — they compare the actual characters or words present in two strings. A paraphrase that conveys the exact same meaning with entirely different words scores poorly on every one of them. Semantic similarity takes a different approach: instead of comparing the strings directly, it compares their *meaning*.

The way this works is through embeddings. A language model reads each string and produces a list of numbers — typically several hundred of them — that collectively encode what the string means. Strings that mean similar things end up with similar numbers; strings that mean different things end up with different numbers. Similarity is then computed by measuring how close two strings' number lists are to each other.

This section uses the `rollama` package, which calls a locally running Ollama instance to generate embeddings. The model is `nomic-embed-text`, a compact embedding model that produces 768-dimensional embeddings. If you have not already pulled it, run `ollama pull nomic-embed-text` in a terminal first.

```{r}
#| label: semantic-setup
#| eval: false
library(rollama)
```

### The concept

To see what embeddings do differently, consider two sentences that say the same thing in completely different words:

```{r}
#| label: semantic-concept
#| eval: false
cosine_sim <- function(a, b) sum(a * b) / (sqrt(sum(a^2)) * sqrt(sum(b^2)))

as_vec <- function(embedding) embedding |> unlist() |> as.numeric()

a <- as_vec(embed_text("the cat sat on the mat",     model = "nomic-embed-text"))
b <- as_vec(embed_text("a feline rested on the rug", model = "nomic-embed-text"))
c <- as_vec(embed_text("the economy grew last quarter", model = "nomic-embed-text"))

cosine_sim(a, b)  # paraphrase: should be high
cosine_sim(a, c)  # unrelated: should be low
```

The paraphrase scores high because the model has learned that "cat" and "feline" refer to the same thing, and that "sat" and "rested" describe the same kind of action. The surface metrics earlier on this page would give both pairs a similarly low score, because they share almost no words.

### Applied to the translations

```{r}
#| label: semantic-translations
#| eval: false
source_vec <- as_vec(embed_text(source, model = "nomic-embed-text"))

translations |>
  mutate(
    embedding  = map(text, \(t) as_vec(embed_text(t, model = "nomic-embed-text"))),
    similarity = map_dbl(embedding, cosine_sim, b = source_vec)
  ) |>
  select(label, similarity)
```

Unlike BLEU and the `stringsim()` metrics, semantic similarity can recognise that the paraphrase translation is closer to the source than the score from surface overlap suggests. The unrelated and different translations should still score low, because their meaning genuinely differs.

### When to use this

Surface metrics are the right default when you expect translations to be close to the reference — same words, maybe slightly reordered or with small substitutions. They are fast, require no external model, and are easy to reason about.

Semantic similarity is worth reaching for when paraphrasing is expected — valid translations that choose different vocabulary or restructure sentences while preserving meaning. In that case, surface metrics will penalise good translations, and embedding-based similarity gives a more accurate picture.

## Language considerations

All five methods on this page were developed primarily with English in mind, and they behave somewhat differently across languages. The table below summarises the situation for Dutch, English, Turkish, Polish, and Arabic specifically.

| Language | Script | Space-delimited | Morphology | Notes |
|---|---|---|---|---|
| English | Latin | Yes | Low | All methods work well |
| Dutch | Latin | Yes | Low–moderate | All methods work well |
| Polish | Latin + diacritics | Yes | Moderate | Works well; diacritics affect character metrics slightly |
| Turkish | Latin | Yes | Very high | Word-level metrics unreliable |
| Arabic | Arabic | Yes | Very high | Word-level metrics unreliable; diacritics add noise |

### Space-delimited languages

All five languages use spaces to separate words, so Jaccard and BLEU tokenisation works correctly for all of them — unlike Chinese or Japanese, where splitting on whitespace does not produce meaningful tokens.

### Morphologically rich languages

Turkish and Arabic are both highly agglutinative or inflected: a single root word can produce dozens of forms depending on tense, case, possession, and other markers. In Turkish, *evlerinden* ("from their houses") is one token derived from *ev* ("house"). In Arabic, the definite article *ال* (al-) is attached directly to the following noun with no space.

This matters most for Jaccard and BLEU, which match exact word tokens. Two sentences that express the same idea in Turkish or Arabic may share very few word-form tokens, producing low scores not because the translations differ but because the same meaning takes different surface forms. Character-level metrics — Levenshtein and cosine — are less affected, because they still pick up shared character sequences within and across word boundaries.

Semantic similarity handles morphological variation best of all, because the embedding model has been trained to map different surface forms of the same concept to similar vectors.

### Arabic diacritics

Arabic text may or may not include diacritical marks (harakat), which indicate vowel sounds. The same word written with and without diacritics will have a non-zero Levenshtein distance and lower cosine similarity even though it is the same word. If your Arabic texts are inconsistently diacritised, strip diacritics during cleaning:

```{r}
#| label: arabic-clean
#| eval: false
strip_arabic_diacritics <- function(x) {
  str_remove_all(x, "[ً-ٰٟ]")
}
```

### Embeddings across languages

`nomic-embed-text` is trained primarily on English and will produce lower-quality embeddings for Turkish and Arabic in particular. For multilingual use, swap it for one of the following models, both available in Ollama and covering all five languages:

- **`nomic-embed-text-v2-moe`** (~1 GB) — Nomic's official multilingual model, trained on over 1.6 billion pairs across ~100 languages. A direct upgrade from `nomic-embed-text` that works with the same `rollama` code.
- **`jeffh/intfloat-multilingual-e5-large-instruct`** (~600 MB quantized) — one of the most widely used multilingual embedding models; the quantized version is smaller than `nomic-embed-text-v2-moe` with comparable quality.

Pull whichever you prefer with `ollama pull <model>` and replace `"nomic-embed-text"` with the model name in the code above. No Python required.

## Sentence-level vs. document-level

The examples above compare single sentences. When working with full documents, you have two options:

**Document-level:** concatenate each translation into one long string and compute similarity once. This is fast but hides variation within a document — a translation that is excellent in one section and poor in another looks the same as one that is mediocre throughout.

**Sentence-level:** align the source and translation sentence by sentence, compute similarity for each pair, then summarize across sentences. This requires aligned data but reveals where translations diverge.

```{r}
#| label: sentence-level
sentences <- tribble(
  ~sentence, ~source,                             ~translation_a,                        ~translation_b,
  1,         "The cat sat on the mat.",            "The cat sat on the mat.",             "A feline rested upon the rug.",
  2,         "It looked out the window.",          "It looked out the window.",            "The animal gazed outside.",
  3,         "The sun was bright that afternoon.", "The sun was bright that afternoon.",   "It was a sunny day in the evening."
) |>
  mutate(across(source:translation_b, clean))

sentences |>
  mutate(
    sim_a = stringsim(translation_a, source, method = "lv"),
    sim_b = stringsim(translation_b, source, method = "lv")
  ) |>
  summarise(
    mean_a = mean(sim_a),
    mean_b = mean(sim_b),
    min_a  = min(sim_a),
    min_b  = min(sim_b)
  )
```

The sentence-level summary shows not just the average similarity but also where a translation is weakest.
