---
title: "Sequential analyses"
toc: true
code-tools: true
code-fold: true
---

```{r}
#| label: setup
#| message: false

# Load packages
library(MASS)
library(tidyverse)
library(viridis)
library(brms)
library(emmeans)
library(BayesFactor)
library(tidybayes)

# Set the default ggplot theme
theme_set(theme_minimal())

# Set seed
set.seed(1)
```

This section is about performing sequential analyses.

## Bayesian

### Two groups

For this scenario we simulate data from two different groups (say a control group and an experimental group).

```{r}
#| label: simulation-2-groups

# Set the simulation parameters
Ms <- c(0, 0.25)
SDs <- 1
n <- 250
labels <- c("control", "experimental")

# Produce the variance-covariance matrix
Sigma <- matrix(
  nrow = length(Ms), 
  ncol = length(Ms), 
  data = c(
    SDs^2, 0,
    0, SDs^2
  )
)

# Simulate
m <- mvrnorm(n = n, mu = Ms, Sigma = Sigma, empirical = TRUE)

# Prepare data
colnames(m) <- labels

data <- as_tibble(m)

data <- pivot_longer(
  data = data, 
  cols = everything(), 
  names_to = "condition", 
  values_to = "DV"
)

data <- mutate(data, id = 1:n(), .before = condition)
```

There are different Bayesian ways to analyze this data. One can focus on estimation or on Bayes factors.

#### Estimation

If estimation is the goal, we run the model (or models) and obtain the posterior distribution of the estimates of interest. Below we run subsequent models with each model using more of the data.

```{r}
#| label: estimates-2-groups
#| message: false
#| results: false
#| cache: true
ns <- c(10, 20, 30, 40, 50, 75, 100, 150, 200, 250)
results <- tibble()

for (i in 1:length(ns)) {
  # Get the sample size
  n <- ns[i]
  
  # Draw a sample of size n
  sample <- slice_head(data, n = n)
  
  # If this is the first iteration, run the full brms model
  # Else update the model
  if (i == 1) {
    model <- brm(
      formula = DV ~ 0 + Intercept + condition, 
      data = sample, 
      family = gaussian(), 
      prior = c(
        set_prior(coef = "Intercept", prior = "normal(0, 1)"),
        set_prior(class = "b", prior = "normal(0, 1)"),
        set_prior(class = "sigma", prior = "normal(1, 1)")
      )
    )
  } else {
    model <- update(model, newdata = sample)
  }
  
  # Get the posteriors of the model estimates
  draws <- as_draws_df(model)
  
  # Add sample information
  draws <- mutate(draws, step = i, n = n)
  
  # Add the draws to the results data frame
  results <- bind_rows(
    results, 
    draws
  )
}
```

Now we have the draws of the posterior distribution of the model estimates. We can plot these using the following code.

```{r}
#| label: estimates-plot-2-groups
#| fig-cap: Sequential analyses of 2 groups
#| fig-subcap: Horizontal lines show the 95% interval of the estimate from the largest sample.
final_quantiles <- results %>%
  filter(n == 250) %>%
  pull(b_conditionexperimental) %>%
  quantile(probs = c(.025, .975))

ggplot(results, aes(x = factor(n), y = b_conditionexperimental)) +
  stat_halfeye() +
  geom_hline(yintercept = final_quantiles, linetype = "dashed", alpha = .5) +
  labs(x = "Sample size per condition (n)", y = "")
```

### Bayes factors

This is still a work in progress. I have yet to figure out the best way to obtain Bayes factors using the `brms` package.

```{r}
#| label: sequential-analyses-2-groups
#| eval: false
n_step <- 50
results <- tibble()

for (i in 1:(n / n_step)) {
  # Draw a sample
  sample <- slice_head(data, n = i * n_step)
  
  # Run the brm model
  model <- brm(
    formula = DV ~ 0 + Intercept + condition, 
    data = sample, 
    family = gaussian(), 
    prior = c(
      set_prior(coef = "Intercept", prior = "normal(0, 1)"),
      set_prior(class = "b", prior = "student_t(1, 0, 1)"),
      #set_prior(class = "sigma", prior = "normal(1, 1)"),
      set_prior("", "sigma") + 
        set_prior("target += log(1 / sigma)", check = FALSE)
    ), 
    sample_prior = TRUE
  )
  
  # Calculate the BF of a contrast
  BF_brms <- hypothesis(model, "conditionexperimental = 0")

  BF_BF <- ttestBF(formula = DV ~ condition, data = sample, rscale = "wide")
  
  # Add the information to the bayes factors data frame
  results <- bind_rows(
    bayes_factors, 
    tibble(
      step = i,
      n = i * n_step,
      BF_brms = BF_brms$hypothesis$Evid.Ratio,
      BF_BF = extractBF(BF_BF)$bf
    )
  )
}
```

### Four groups

Let's simulate data for a scenario in which we have 4 different between-subjects conditions. The conditions differ from each by a small amount and for simplicity's sake each condition has a standard deviation of 1.

```{r}
#| label: simulation-4-groups

# Set the simulation parameters
Ms <- c(0, 0.2, 0.4, 0.6)
SDs <- 1
n <- 250
labels <- c("A", "B", "C", "D")

# Produce the variance-covariance matrix
Sigma <- matrix(
  nrow = length(Ms), 
  ncol = length(Ms), 
  data = c(
    SDs^2, 0, 0, 0,
    0, SDs^2, 0, 0,
    0, 0, SDs^2, 0,
    0, 0, 0, SDs^2
  )
)

# Simulate
m <- mvrnorm(n = n, mu = Ms, Sigma = Sigma, empirical = TRUE)

# Prepare data
colnames(m) <- labels

data <- as_tibble(m)

data <- pivot_longer(
  data = data, 
  cols = everything(), 
  names_to = "condition", 
  values_to = "DV"
)

data <- mutate(data, id = 1:n(), .before = condition)
```

With this data we can run multiple sequential models (like in the 2 groups scenario), except this time we calculate contrasts between all the levels of the condition factor. We again obtain the posteriors of these contrasts and store them so we can plot them afterwards.

```{r}
#| label: sequential-analyses-4-groups
#| message: false
#| results: false
#| cache: true
ns <- c(10, 20, 30, 40, 50, 75, 100, 150, 200, 250)
results <- tibble()

for (i in 1:length(ns)) {
  # Get the sample size
  n <- ns[i]
  
  # Draw a sample of size n
  sample <- slice_head(data, n = n)
  
  # If this is the first iteration, run the full brms model
  # Else update the model
  if (i == 1) {
    model <- brm(
      formula = DV ~ 0 + Intercept + condition, 
      data = sample, 
      family = gaussian(), 
      prior = c(
        set_prior(coef = "Intercept", prior = "normal(0, 1)"),
        set_prior(class = "b", prior = "normal(0, 1)"),
        set_prior(class = "sigma", prior = "normal(1, 1)")
      )
    )
  } else {
    model <- update(model, newdata = sample)
  }
  
  # Get the estimated marginal means
  emmeans <- emmeans(model, specs = pairwise ~ condition)
  contrasts <- emmeans$contrasts
  
  # Get draws of the posterior of each contrast
  draws <- gather_emmeans_draws(contrasts)
  
  # Add sample information
  draws <- mutate(draws, step = i, n = n)
  
  # Add the draws to the results data frame
  results <- bind_rows(
    results, 
    draws
  )
}
```

Now that we have a data frame that contains the posterior draws of each contrast, we can plot the posteriors as well as some summary statistics (e.g., the median, a 95% interval) for each contrast.

```{r}
#| label: plot-4-groups
#| fig-cap: Sequential analysis
#| fig-height: 14
final_quantiles <- results %>%
  filter(n == 250) %>%
  group_by(contrast) %>%
  summarize(
    final_lower = quantile(.value, .025),
    final_upper = quantile(.value, .975)
  ) %>%
  pivot_longer(cols = -contrast, names_to = "bound", "value")

ggplot(results, aes(x = factor(n), y = .value)) +
  stat_slabinterval() +
  geom_hline(
    mapping = aes(yintercept = value), 
    data = final_quantiles, 
    linetype = "dashed", 
    alpha = .5
  ) +  
  facet_wrap(~ contrast, ncol = 1, scales = "free_y") +
  labs(x = "Sample size per condition (n)", y = "Contrast estimate") +
  scale_color_viridis(option = "mako", discrete = TRUE)
```

