---
title: "Speciesism Scale"
execute:
  echo: false
format: 
   html:
     df-print: kable
     toc: true
knitr:
  opts_chunk: 
    R.options:
      knitr.kable.NA: ''
nocite: |
  @caviola2019
bibliography: references.bib
---

```{r}
#| label: setup
#| message: false
library(tidyverse)
library(DT)
items <- read_csv("speciesism-items.csv")
```

The Speciesism Scale is a 6-item measure of speciesism---the assignment of different inherent moral status based solely on an individual's species membership.

## Items

```{r}
#| results: asis
items %>%
  select(item_nr, item) %>%
  unite(col = item, everything(), sep = ". ") %>%
  pull(item) %>%
  str_c("\n") %>%
  str_flatten() %>%
  cat()
```

## Response options

-   Strongly disagree
-   Disagree
-   Somewhat disagree
-   Neither agree nor disagree
-   Somewhat agree
-   Agree
-   Strongly agree

::: callout-note
The original paper only reports the labels for the end points of the scale ('Strongly disagree' and 'Strongly agree'), not the options in between.
:::

## Download

```{r}
#| label: table
#| tbl-cap: Speciesism Scale items

items %>%
  select(where(~ sum(is.na(.x)) != length(.x))) %>%
  datatable(
    rownames = FALSE, 
    extensions = 'Buttons', 
    options = list(
      dom = 'Bip',
      buttons = list(
        'copy', 
        list(
          extend = 'collection',
          buttons = c('csv', 'excel'),
          text = 'Download'
        )
      )
    )
  )
```

## Reference

::: {#refs}
:::