---
title: "Family-Wise Error"
description: "What family-wise error is, how the claim being tested determines whether it applies, and when correction is worth the loss in power."
toc: true
draft: true
---

```{r}
#| label: setup
#| message: false
#| code-fold: true
library(tidyverse)

theme_set(theme_minimal())

color_primary <- "#2171b5"
color_secondary <- "#888888"
color_reference <- "gray50"
```

## What family-wise error is

A single hypothesis test run at α = .05 has a 5% chance of a false positive when the null hypothesis is true. Running more than one test on the same data raises the probability that at least one of them turns up significant by chance alone, even though each test still runs at α = .05. The family-wise error rate is the probability of at least one false positive across a set, or family, of tests, all conducted under a true null.

Whether the family-wise error rate is something to worry about depends on the claim the tests are being used to support.

## Claim 1: At least one of several tests shows a real effect

The first scenario is a claim that holds as soon as any one of several tests comes out significant. Such a claim arises whenever a researcher runs a batch of tests and treats a single significant result as enough to support it. Comparing a treatment group to a control group on several outcome measures is one example: not knowing in advance which outcome the treatment might affect, the researcher claims that it affects at least one of them. Computing a batch of correlations and claiming that at least one of the variables is related to the outcome is another, as is testing several subgroups and claiming an effect in at least one. In each case the claim is disjunctive: any single significant test is taken to support it. This is the structure behind much of what is called researcher degrees of freedom {{< cite simmons2011 >}}. Measuring several outcomes, splitting the sample several ways, or running several analyses and keeping whichever comparison reached significance all give the disjunctive claim more chances to be wrongly supported.

Suppose that the claim is false: none of the tested effects is real, so every test is a true null. The claim is wrongly supported whenever at least one test is significant by chance, and the probability of that grows as more tests are run. The graph below shows this rate climbing with the number of tests, alongside the same rate once a Bonferroni correction is applied.

```{r}
#| label: fwer-any-effect
#| code-fold: true
#| fig-cap: "Probability of wrongly claiming at least one real effect, without correction and under a Bonferroni correction, as a function of the number of tests run. The dashed line marks α = .05."

alpha <- 0.05
k_max <- 20

fwer_any_effect <- tibble(
  m = 1:k_max,
  Uncorrected = 1 - (1 - alpha)^m,
  Bonferroni = 1 - (1 - alpha / m)^m
) |>
  pivot_longer(
    c(Uncorrected, Bonferroni),
    names_to = "method",
    values_to = "fwer"
  )

ggplot(fwer_any_effect, aes(x = m, y = fwer, color = method)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = alpha, linetype = "dashed", color = color_reference) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_color_manual(
    values = c("Uncorrected" = color_primary, "Bonferroni" = color_secondary)
  ) +
  labs(
    x = "Number of tests run (m)",
    y = "P(claim wrongly asserted)",
    color = NULL
  )
```

Without correction, the false-claim rate is 1 - (1 - α)^m. With a single test it equals α, the per-comparison rate. With five tests it has already climbed past 20%, and with twenty, past 60%.

A Bonferroni correction tests each of the m comparisons at α / m instead of α. The false-claim rate is then 1 - (1 - α/m)^m, which stays just under α no matter how many tests are run, the flat line along the dashed reference. Dividing the threshold by the number of tests holds the family-wise error rate at the intended level.

The claim here is disjunctive: it is true as soon as any one of the tested effects is real, so a false claim requires only one test to be a false positive. That probability is exactly the family-wise error rate, which is why this is the kind of claim a correction is built for.

## Claim 2: Every one of several tests shows a real effect

The second scenario is the conjunctive counterpart to the first: where the disjunctive claim held as soon as one test was significant, this one holds only if every test is significant. Comparing four treatment conditions against a shared control and claiming that all four beat it is one example. Running several correlations and claiming that every one of them is nonzero is another. Mediation has the same structure, since claiming that X affects the mediator, that the mediator affects the outcome, and that the indirect effect is present requires each of those tests to succeed. So does a multi-study paper that claims an effect appears in every one of its studies rather than merely somewhere among them. In each case, asserting the claim means requiring all of its tests to be significant at once.

What counts as a Type I error changes with the claim. Here the claim is false as soon as even one of its components is truly null, so wrongly asserting it means concluding that every effect is real when at least one is not. The error is judged against the whole claim, not the individual tests. To make it, each truly null test has to reach significance by chance while every genuine test also comes out significant. That hinges on the null test firing by chance, an event with probability α, so the error rate cannot exceed α. With more than one null test, several chance hits would have to line up at once, which only makes it smaller.

The graph below traces this maximum. It takes the worst case, where exactly one of the tests is a true null and the rest carry effects large enough to be detected essentially every time. The full claim is then wrongly asserted exactly when that single null test fires by chance, so the worst-case rate depends on that one test alone, not on how many genuine tests sit alongside it. The graph plots this rate against the number of tests in the conjunction, without correction and under a Bonferroni correction that divides α by the number of tests.

```{r}
#| label: fwer-all-effects
#| code-fold: true
#| fig-cap: "Maximum probability of wrongly asserting the full conjunction, without correction and under a Bonferroni correction, as a function of the number of tests in the conjunction. The dashed line marks α = .05."

alpha <- 0.05
k_max <- 20

fwer_all_effects <- tibble(
  m = 1:k_max,
  Uncorrected = alpha,
  Bonferroni = alpha / m
) |>
  pivot_longer(
    c(Uncorrected, Bonferroni),
    names_to = "method",
    values_to = "fwer"
  )

ggplot(fwer_all_effects, aes(x = m, y = fwer, color = method)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = alpha, linetype = "dashed", color = color_reference) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_color_manual(
    values = c("Uncorrected" = color_primary, "Bonferroni" = color_secondary)
  ) +
  labs(
    x = "Number of tests in the conjunction (m)",
    y = "Maximum P(claim wrongly asserted)",
    color = NULL
  )
```

Without correction, the worst-case false-claim rate sits at α and stays there as more tests are added to the conjunction. This is the mirror image of the first scenario. There, each additional test gave the disjunctive claim another way to fail, so the error rate climbed with the number of tests. Here, each additional test is one more result that must come out significant, which can only hold the error rate down. The uncorrected conjunction is already a valid test at level α no matter how many tests it contains.

Bonferroni pushes the rate down to α divided by the number of tests, well below the α it was already at, so it buys no protection the claim needed. The cost shows up when the claim is true and every effect is real: the stricter α/m threshold on every test lowers the chance of correctly asserting the whole conjunction, which is the power to detect the effect.

A claim that holds only when every one of several tests holds is called an intersection-union test. Testing each component at the uncorrected level α already keeps the false-claim rate at or below α, regardless of how many tests are involved or how correlated they are. Correcting for multiplicity here trades power for protection the claim never needed.

## Other kinds of claim

Not every research claim is a clean disjunction or conjunction. Two common ones fall outside the pair above and call for a different treatment.

The first is a claim about which specific effects are real. A researcher who tests many effects and reports the subset that came out significant is not asserting one claim but making a separate decision about each effect, so there is no single claim to be true or false. The family-wise error rate is not the right target here. What matters instead is the false discovery rate, the expected fraction of the reported set that is a false positive, and it calls for its own kind of correction, such as Benjamini-Hochberg.

The second is a claim about the pattern the results form. A moderation claim is the common example, often stated as an effect being present in one condition and absent in another. It is neither disjunctive nor conjunctive, because it mixes a rejection with a non-rejection: showing the effect in the first condition means rejecting a null, but showing its absence in the second means affirming one, and a non-significant test is not evidence that the effect is zero.

The claim that the two conditions differ from each other is a single test, the interaction, and that is what should carry the moderation claim in the first place. Comparing a significant simple effect in one condition against a non-significant one in the other is not a test of moderation, because the difference between significant and not significant is not itself significant {{< cite gelman2006 >}}. Even a significant interaction does not settle the matter, since it shows only that the effect differs across conditions, not that it follows the specific pattern that was predicted. Recovering that pattern calls for follow-up tests of the simple effects, which bring back the same multiplicity questions for a claim whose logic is harder to pin down.

## Conclusion

Whether the family-wise error rate is worth worrying about comes down to the claim the tests support, not the number of tests. A disjunctive claim, true as soon as one test is significant, is put at risk by every extra test and needs a correction. A conjunctive claim, true only if every test is significant, is protected by its own structure and needs none. A single planned comparison is a family of one, where the family-wise error rate is already α and there is nothing to correct.

Applying this means deciding which tests belong to the same family, which is a judgment call rather than something the data can settle. A family is the set of tests that support one claim, so two researchers can look at the same analyses and disagree about how many families they contain, because the answer depends on what each test is being used to claim.

Where a correction is warranted, it is a trade rather than a free improvement. Holding the family-wise error rate at α means a stricter threshold on each test, which lowers the power to detect a real effect. Whether the trade is worth it depends on the cost of a false positive against the cost of missing a true effect: a confirmatory result feeding a costly decision calls for tighter control, while an exploratory scan meant to surface candidates for later follow-up can accept a higher error rate for more power.

The [omnibus F-test](../omnibus-test/) page works through this trade-off concretely for pairwise comparisons following an ANOVA, comparing an uncorrected procedure against Holm and Tukey correction under a partial null.
