---
title: "Animal Attitude Scale"
execute:
  echo: false
knitr:
  opts_chunk: 
    R.options:
      knitr.kable.NA: ''
nocite: |
  @herzog1991, @herzog2015
---

```{r}
#| label: setup
#| message: false
library(tidyverse)
items <- read_csv("AAS-items.csv")
```

The Animal Attitudes Scale is a 20-item measure of pro-animal welfare attitudes. Two short versions of the scale are available (a 5 and a 10 item version).

## Items

```{r}
#| results: asis
items %>%
  select(item_nr, item) %>%
  unite(col = item, everything(), sep = ". ") %>%
  pull(item) %>%
  str_c("\n") %>%
  str_flatten() %>%
  cat()
```

## Instructions

> Listed below are 20 statements regarding the use of animals. Circle the letters that indicate the extent to which you agree or disagree with the statement.

## Response options

-   Strongly disagree
-   Disagree
-   Undecided
-   Agree
-   Strongly agree

## Scale items

```{r}
#| label: table

items %>%
  select(where(~ sum(is.na(.x)) != length(.x))) %>%
  knitr::kable()
```

## Reference

::: {#refs}
:::