---
title: "Oxford Utilitarianism Scale"
execute:
  echo: false
format: 
   html:
     df-print: kable
     toc: true
knitr:
  opts_chunk: 
    R.options:
      knitr.kable.NA: ''
nocite: |
  @kahane2018
bibliography: references.bib
---

```{r}
#| label: setup
#| message: false
library(tidyverse)
library(DT)
items <- read_csv("OUS-items.csv")
```

The scale consists of 9 items in two subscales. The first subscale--Impartial Beneficence (OUSIB)--consists of 5 items that all tap endorsement of the impartial maximization of the greater good, even at the cost of personal self-sacrifice. The second subscale was labeled Instrumental Harm (OUS-IH). This subscale consists of 4 items that all tap into a willingness to cause harm to bring about the greater good.

## Items

```{r}
#| results: asis
items %>%
  select(item_nr, item) %>%
  unite(col = item, everything(), sep = ". ") %>%
  pull(item) %>%
  str_c("\n") %>%
  str_flatten() %>%
  cat()
```

## Instructions

> Indicate how much you agree or disagree with each of the following statements (1 = strongly disagree, 4 = neither agree nor disagree, 7 = strongly agree)

## Response options

A 7-point Likert scale with at least 3 of the response options labeled: 1 = strongly disagree, 4 = neither agree nor disagree, 7 = strongly agree.

## Download

```{r}
#| label: table
#| tbl-cap: Speciesism Scale items

items %>%
  select(where(~ sum(is.na(.x)) != length(.x))) %>%
  datatable(
    rownames = FALSE,
    extensions = "Buttons",
    options = list(
      dom = "Bip",
      buttons = list(
        "copy",
        list(
          extend = "collection",
          buttons = c("csv", "excel"),
          text = "Download"
        )
      )
    )
  )
```

## Reference

::: {#refs}
:::