---
title: "Machine Translation"
description: "When machine translation can stand in for human survey-translation protocols, used with care"
toc: true
---

```{r}
#| label: setup
#| message: false
#| code-fold: true
library(tidyverse)
library(deeplr)
library(googleLanguageR)
library(ellmer)
```

Survey translation has traditionally required human translators working through structured protocols like TRAPD (Translation, Review, Adjudication, Pretesting, Documentation). MT systems have become accurate enough to serve as a first pass or, in some contexts, a full replacement — but they require some care when applied to survey text.

Survey items present specific challenges for MT. They are short, often grammatically incomplete, and highly decontextualised: a response option like "Agree strongly" or an interviewer instruction like "Read out" means nothing without knowing what question it belongs to. Terminology also needs to be consistent across items — the same concept should always be translated the same way throughout a questionnaire. Both of these problems are addressed by providing context to the MT system.

## Context

MT systems trained on sentence pairs translate each sentence independently by default. For general text this is usually fine, but for surveys it causes two problems:

- **Ambiguity**: short items are more ambiguous than full sentences, and the correct translation often depends on the surrounding questionnaire. "How often?" could translate differently depending on whether the preceding item asked about behaviour, attitudes, or frequency of contact.
- **Inconsistency**: without seeing the full questionnaire, an MT system may translate the same term differently across items — using two different words for the same concept in the same questionnaire.

Providing the surrounding questionnaire as context mitigates both. How you do this depends on the MT system.

## MT systems

### DeepL

DeepL is among the most widely used MT systems in translation research and produces high-quality output for European languages. It offers three ways to incorporate context.

**Automatic document context** is the default when you pass multiple sentences at once. DeepL processes surrounding sentences when translating each one, so simply passing several items together improves coherence compared to passing them one at a time.

**The `context` parameter** is the cleanest approach for item-by-item translation. You pass the text you want translated as `text` and the surrounding questionnaire as `context`. DeepL uses the context to inform the translation but only returns the translation of `text` — there is no extraction problem.

```{r}
#| label: deepl-context
#| eval: false
translate2(
  text        = "Generally speaking, would you say that most people can be trusted?",
  target_lang = "NL",
  context     = "This item is part of a social attitudes survey.
                 The response options are: Most people can be trusted;
                 Can't be too careful; Don't know.",
  auth_key    = Sys.getenv("DEEPL_API_KEY")
)
```

To translate all items in a questionnaire with their surrounding context:

```{r}
#| label: deepl-context-all
#| eval: false
questionnaire <- tribble(
  ~item_id, ~text,
  1, "How satisfied are you with your life as a whole?",
  2, "Generally speaking, would you say that most people can be trusted?",
  3, "How interested would you say you are in politics?"
)

full_context <- paste(questionnaire$text, collapse = "\n")

questionnaire |>
  mutate(
    translation = map_chr(text, \(item) {
      translate2(
        text        = item,
        target_lang = "NL",
        context     = full_context,
        auth_key    = Sys.getenv("DEEPL_API_KEY")
      )
    })
  )
```

**The document endpoint** accepts a full `.docx` or `.txt` file and translates it as a whole, using the entire document as its own context. This is convenient for full questionnaires, but you receive the full translated document back and need to parse individual items out afterwards. The `context` parameter approach avoids this and is generally preferable for structured questionnaire data.

### Google Translate

Google Translate has no equivalent to DeepL's `context` parameter — you cannot explicitly pass surrounding text to inform the translation of a single item. It translates each piece of text you give it without awareness of other items in the questionnaire unless you pass them together.

What it does offer, via the v3 (Advanced) API, is **glossary support**: you define a set of source→target term pairs and the API enforces those translations consistently across all requests. This addresses the terminology consistency problem directly, even without sentence-level context.

```{r}
#| label: google-glossary
#| eval: false
library(googleLanguageR)

# Glossaries are created once via the Cloud Translation API and referenced by ID.
# Here we use an existing glossary and translate a batch of items.
gl_translate(
  string      = questionnaire$text,
  target      = "nl",
  source      = "en",
  glossaryId  = "my-survey-glossary",
  key         = Sys.getenv("GOOGLE_API_KEY")
)
```

One limitation: glossary support requires the Cloud Translation v3 API, which is billed differently from the basic v2 API. The `googleLanguageR` package supports both.

### LLMs

LLM-based translation puts the entire instruction in a prompt. The text to translate, the surrounding questionnaire, a list of required term translations, and notes on register or target population all go in the same place. This makes an LLM the most flexible of the three systems described here. The flexibility comes with a cost: a prompt that only says "translate this" supplies no more than a sentence-level MT system has, so it reproduces the same failures. Improving an LLM translation means adding the missing information to the prompt. The sections below first show what a minimal prompt gets wrong, then fix each problem by adding to the prompt.

The examples translate Dutch survey text into English using a local model served through LM Studio and called with the `ellmer` package. The `chat_lmstudio()` function opens a connection to the local model, `system_prompt` sets the standing instruction, and `params(temperature = 0)` makes repeated runs return the same output.

#### A bare prompt reproduces sentence-level failures

The minimal prompt translates each piece of text on its own, with no information about the questionnaire it belongs to. That is usually enough for a full sentence, but survey response options are short, and a short Dutch word that is unambiguous inside a sentence can carry more than one meaning on its own. Take the response option "Eens". On a rating scale it means "agree", but in isolation it also reads as "once", and translated without the question it answers it comes back as "once", the wrong reading for a survey. What would settle the meaning is the question the option belongs to, and that is exactly what the minimal prompt leaves out.

#### Supplying context

Supplying context takes two changes to the call. The instruction has to state that the text is a survey response option and tell the model to read it in light of the question, and the question itself has to be included in the input. The chunk below keeps the original bare prompt as `translate_without_context`, adds `translate_with_context` for the version that makes both changes, and runs each on the same option.

```{r}
#| label: llm-context-fix
#| eval: false
translate_without_context <- function(text) {
  chat <- chat_lmstudio(
    model         = "google/gemma-4-26b-a4b-qat",
    system_prompt = "Translate the following text from Dutch to English. Output only the translation.",
    params        = params(temperature = 0)
  )
  chat$chat(text)
}

translate_with_context <- function(option, question) {
  chat <- chat_lmstudio(
    model         = "google/gemma-4-26b-a4b-qat",
    system_prompt = paste0(
      "Translate the Dutch survey response option to English. ",
      "Use the survey question for context. Output only the translation."
    ),
    params        = params(temperature = 0)
  )
  chat$chat(paste0("Question: ", question, "\nResponse option: ", option))
}

tibble(
  option          = "Eens",
  without_context = translate_without_context("Eens"),
  with_context    = translate_with_context(
    "Eens",
    "In hoeverre bent u het eens met de volgende stelling?"
  )
)
```

|option |without_context |with_context |
|:------|:---------------|:------------|
|Eens   |Once            |Agree        |

#### Supplying a glossary

A glossary is a list of required translations for specific terms. It is the right tool when a term has no clean English equivalent and the project has agreed on a standard rendering, so that every occurrence is translated the same way across items and across survey rounds. It is not a tool for resolving meaning: a word whose correct sense depends on the sentence, like the response options above, belongs in context rather than pinned to a fixed translation.

The agreed renderings go into the prompt as explicit term pairs.

```{r}
#| label: llm-glossary-fix
#| eval: false
edu_items <- tibble(dutch = c("VWO", "HAVO", "VMBO"))

glossary <- c(
  "VWO  = pre-university education (VWO)",
  "HAVO = senior general secondary education (HAVO)",
  "VMBO = pre-vocational secondary education (VMBO)"
)

translate_with_glossary <- function(text) {
  chat <- chat_lmstudio(
    model         = "google/gemma-4-26b-a4b-qat",
    system_prompt = paste0(
      "Translate the following Dutch text to English.\n",
      "Use these required translations for specific terms:\n",
      paste(glossary, collapse = "\n"), "\n",
      "Output only the translation."
    ),
    params        = params(temperature = 0)
  )
  chat$chat(text)
}

edu_items |>
  mutate(
    without_glossary = map_chr(dutch, translate_without_context),
    with_glossary    = map_chr(dutch, translate_with_glossary)
  )
```

|dutch |without_glossary                          |with_glossary                             |
|:-----|:-----------------------------------------|:-----------------------------------------|
|VWO   |Pre-university education                  |pre-university education (VWO)            |
|HAVO  |HAVO (Higher General Continued Education) |senior general secondary education (HAVO) |
|VMBO  |Pre-vocational secondary education        |pre-vocational secondary education (VMBO) |

The same glossary applies when an education level appears inside a full item rather than on its own. The rest of the sentence is translated normally, and the listed term takes its agreed rendering.

```{r}
#| label: llm-glossary-sentence
#| eval: false
item <- "Bent u na de HAVO doorgestroomd naar het hoger onderwijs?"

tibble(
  dutch            = item,
  without_glossary = translate_without_context(item),
  with_glossary    = translate_with_glossary(item)
)
```

|dutch                                                     |without_glossary                                |with_glossary                                                                        |
|:---------------------------------------------------------|:-----------------------------------------------|:------------------------------------------------------------------------------------|
|Bent u na de HAVO doorgestroomd naar het hoger onderwijs? |Did you proceed to higher education after HAVO? |Did you proceed to higher education after senior general secondary education (HAVO)? |

#### Combining context and a glossary

Translating a real questionnaire combines both techniques, but the context an item needs is specific: a response option needs its own question, not the entire questionnaire. Recording the survey structure makes that possible. Each row stores the text, what kind of element it is, the question it belongs to, and the block it sits in. A block can span several questions, so `question` identifies the question and `block` groups related questions together.

```{r}
#| label: llm-combined
#| eval: false
questionnaire <- tribble(
  ~block,      ~question, ~type,      ~text,
  "agreement", 1,         "question", "In hoeverre bent u het eens met de volgende stelling?",
  "agreement", 1,         "option",   "Eens",
  "education", 2,         "question", "Wat is uw hoogst voltooide opleiding?",
  "education", 2,         "option",   "VWO",
  "education", 2,         "option",   "HAVO",
  "education", 2,         "option",   "VMBO",
  "education", 3,         "question", "Bent u na de HAVO doorgestroomd naar het hoger onderwijs?"
)
```

The prompt for each item is built from these columns. It names the element type, and for a response option it supplies that option's own question as context, found by matching on `question`. The glossary is added as before.

```{r}
#| label: llm-combined-translate
#| eval: false
stems <- questionnaire |>
  filter(type == "question") |>
  select(question, stem = text)

translate_item <- function(text, type, block, stem) {
  type_desc <- switch(type,
    question    = "a survey question",
    option      = "a response option",
    instruction = "interviewer instruction text"
  )

  system_prompt <- paste0(
    "Translate Dutch survey text to English.\n",
    "The text is ", type_desc, " from the \"", block, "\" block of a survey.\n",
    if (type == "option") paste0("It is a response option to the question: \"", stem, "\"\n"),
    "Use these required translations for specific terms:\n",
    paste(glossary, collapse = "\n"), "\n",
    "Output only the translation of the text."
  )

  chat <- chat_lmstudio(
    model         = "google/gemma-4-26b-a4b-qat",
    system_prompt = system_prompt,
    params        = params(temperature = 0)
  )
  chat$chat(text)
}

questionnaire |>
  left_join(stems, by = "question") |>
  mutate(translation = pmap_chr(list(text, type, block, stem), translate_item)) |> kable()
```

|block     | question|type     |text                                                      |stem                                                      |translation                                                                          |
|:---------|--------:|:--------|:---------------------------------------------------------|:---------------------------------------------------------|:------------------------------------------------------------------------------------|
|agreement |        1|question |In hoeverre bent u het eens met de volgende stelling?     |In hoeverre bent u het eens met de volgende stelling?     |To what extent do you agree with the following statement?                            |
|agreement |        1|option   |Eens                                                      |In hoeverre bent u het eens met de volgende stelling?     |Agree                                                                                |
|education |        2|question |Wat is uw hoogst voltooide opleiding?                     |Wat is uw hoogst voltooide opleiding?                     |What is your highest level of completed education?                                   |
|education |        2|option   |VWO                                                       |Wat is uw hoogst voltooide opleiding?                     |pre-university education (VWO)                                                       |
|education |        2|option   |HAVO                                                      |Wat is uw hoogst voltooide opleiding?                     |senior general secondary education (HAVO)                                            |
|education |        2|option   |VMBO                                                      |Wat is uw hoogst voltooide opleiding?                     |pre-vocational secondary education (VMBO)                                            |
|education |        3|question |Bent u na de HAVO doorgestroomd naar het hoger onderwijs? |Bent u na de HAVO doorgestroomd naar het hoger onderwijs? |Did you proceed to higher education after senior general secondary education (HAVO)? |

## Practical recommendation

DeepL's `context` parameter is the most direct way to pass surrounding questionnaire text when translating item by item. Google Translate's glossary support is the most direct way to enforce consistent terminology across all items. LLMs combine both capabilities in a single prompt, and also allow instructions about register, cultural adaptation, and formatting — at the cost of more setup and less predictable output format.

For workflows where terminology consistency across rounds matters — for example, translating a new round of the ESS to match previous rounds — including reference translations from the MCSQ in the prompt encodes that requirement directly.
