---
title: "Formatting Features"
description: "A reference for the citation, callout, code, and math features available when writing pages."
toc: true
order: 1
---

This page is a reference for the formatting features available when writing pages on this site. Each section shows the source syntax you write and the result it produces, so you can copy a pattern directly. The examples below are live: the citations open the reference panel, the callouts are styled, and the figure is rendered from executed code.

## Citations

Citations are backed by a Zotero-managed bibliography. Every citation is clickable and opens a side panel showing the full reference, and optionally one or more supporting quotes.

The simplest form is a plain `@key`, which renders the author and year:

```markdown
Large language models can stand in for survey respondents [@argyle2023].
```

Large language models can stand in for survey respondents [@argyle2023].

Several keys can be grouped in one set of brackets, separated by semicolons:

```markdown
This idea has been examined from several angles [@argyle2023; @bisbee2024].
```

This idea has been examined from several angles [@argyle2023; @bisbee2024].

### Attaching a supporting quote

To pull a quote from the source into the panel, use the `cite` shortcode. The first argument is the key; any string arguments after it are quotes:

```markdown
{{< cite argyle2023 "Supporting quote from the cited work goes here." >}}
```

The quote text shown in these examples is an illustrative placeholder, not a real excerpt. In practice you paste the actual passage. It renders as a normal citation: {{< cite argyle2023 "Supporting quote from the cited work goes here." >}}. Click it to see the quote and reference.

A bare integer as the last argument is treated as a page number, and multiple quote strings are shown as a list:

```markdown
{{< cite brislin1970 "First supporting quote." "Second supporting quote." 185 >}}
```

{{< cite brislin1970 "First supporting quote." "Second supporting quote." 185 >}}

### Grouped citations with per-citation quotes

When several citations share one parenthetical but each needs its own quote, use the `cites` shortcode. Each key is followed by an optional quote string; omit the string for keys that need no quote:

```markdown
{{< cites argyle2023 "Quote for the first source." bisbee2024 "Quote for the second source." >}}
```

{{< cites argyle2023 "Quote for the first source." bisbee2024 "Quote for the second source." >}}

Each author in the group is individually clickable, and the panel shows only that citation's reference and quote. Use `cites` instead of the plain `[@a; @b]` form whenever any citation in the group needs a quote.

## Callouts

Callouts highlight a piece of text in a labelled box. They use the GitHub alert syntax: a blockquote whose first line is `[!TYPE]`. Five types are available.

```markdown
> [!NOTE]
> Useful information the reader should notice.
```

> [!NOTE]
> Useful information the reader should notice.

> [!TIP]
> A helpful suggestion or shortcut.

> [!IMPORTANT]
> Something the reader must not miss.

> [!WARNING]
> A caution about a likely mistake.

> [!CAUTION]
> A risk with serious consequences.

## Code

Fenced code blocks are syntax-highlighted by language. A block tagged `r` is shown but not run, which is the right choice when you want to display code without executing it:

````markdown
```r
model <- lm(y ~ x, data = data)
summary(model)
```
````

```r
model <- lm(y ~ x, data = data)
summary(model)
```

On pages authored in Quarto, a chunk written with `{r}` instead is executed, and its output is embedded below the code:

```{r}
#| label: executed-chunk
x <- rnorm(100, mean = 50, sd = 10)
summary(x)
```

## Math

Mathematical notation is written in LaTeX and rendered with KaTeX. Inline math is wrapped in single dollar signs, so $\bar{x} = \frac{1}{n}\sum_{i=1}^{n} x_i$ sits in the flow of the text.

Display math is wrapped in double dollar signs and is set on its own line:

```markdown
$$
s^2 = \frac{1}{n-1}\sum_{i=1}^{n}(x_i - \bar{x})^2
$$
```

$$
s^2 = \frac{1}{n-1}\sum_{i=1}^{n}(x_i - \bar{x})^2
$$

## Figures and cross-references

A figure produced by an executed chunk can be given a label and a caption. Quarto numbers the figure and lets you refer to it from the prose.

```{r}
#| label: fig-demo
#| fig-cap: "A figure rendered from executed R code."
#| message: false
#| code-fold: true
library(tidyverse)

tibble(x = 1:100, y = cumsum(rnorm(100))) |>
  ggplot(aes(x, y)) +
  geom_line() +
  theme_minimal()
```

Referencing the chunk's label with `@fig-demo` produces a numbered link: @fig-demo shows the result. Figure and table cross-references work this way; section cross-references (`@sec-`) are a known limitation and do not navigate.

## Tables

A data frame returned from an executed chunk is printed as a table:

```{r}
#| label: table-demo
tibble(
  measure = c("Mean", "SD", "Min", "Max"),
  value = c(mean(x), sd(x), min(x), max(x))
)
```

Markdown tables written by hand are also supported:

```markdown
| Column A | Column B |
| -------- | -------- |
| Cell 1   | Cell 2   |
```

| Column A | Column B |
| -------- | -------- |
| Cell 1   | Cell 2   |

## Frontmatter options

Each page begins with a YAML frontmatter block. The fields that affect rendering are:

```yaml
---
title: "Page Title"
description: "One-line summary shown on the home page card."
toc: true # show the right-side table of contents
toc-depth: 2 # deepest heading level included in the TOC
order: 1 # position within its navigation group
---
```
