---
title: "BFI-2-XS"
execute:
  echo: false
format: 
   html:
     df-print: kable
     toc: true
knitr:
  opts_chunk: 
    R.options:
      knitr.kable.NA: ''
nocite: |
  @soto2017
bibliography: references.bib
---

```{r}
#| label: setup
#| message: false
library(tidyverse)
library(DT)
items <- read_csv("BFI-2-XS-items.csv")
```

The BFI-2-XS is a 15-item extra-short form, of the BFI-2. It is are appropriate for research contexts in which, due to pressing concerns about assessment time or respondent fatigue, administering the full BFI-2 would not be feasible. For most studies, however, it is recommended to administer the full measure due to its greater reliability and validity.

## Items

```{r}
#| results: asis
items %>%
  select(item_nr, item) %>%
  unite(col = item, everything(), sep = ". ") %>%
  pull(item) %>%
  str_c("\n") %>%
  str_flatten() %>%
  cat()
```

## Instructions

> Here are a number of characteristics that may or may not apply to you. For example, do you agree that you are someone who likes to spend time with others? Please write a number next to each statement to indicate the extent to which you agree or disagree with that statement.

## Response options

-   Disagree strongly
-   Disagree a little
-   Neutral; no opinion
-   Agree a little
-   Agree strongly

## Download

```{r}
#| label: table
#| tbl-cap: BFI-2-XS items

items %>%
  select(where(~ sum(is.na(.x)) != length(.x))) %>%
  datatable(
    rownames = FALSE,
    extensions = "Buttons",
    options = list(
      dom = "Bip",
      buttons = list(
        "copy",
        list(
          extend = "collection",
          buttons = c("csv", "excel"),
          text = "Download"
        )
      )
    )
  )
```

## Extra

See also this [link](https://www.colby.edu/psych/personality-lab/) for more information, such as versions of different lengths and translations.

## Reference

::: {#refs}
:::