---
title: "TIPI"
execute:
  echo: false
format: 
   html:
     df-print: kable
     css: styles.css
     toc: true
knitr:
  opts_chunk: 
    R.options:
      knitr.kable.NA: ''
bibliography: references.bib
nocite: |
  @gosling2003
---

```{r}
#| label: setup
#| message: false
library(tidyverse)
library(DT)
items <- read_csv("TIPI-items.csv")
```

The TIPI is a 10-item measure of the Big Five (or Five-Factor Model) dimensions.

## Items

```{r}
#| results: asis
items %>%
  select(item_nr, item) %>%
  unite(col = item, everything(), sep = ". ") %>%
  pull(item) %>%
  str_c("\n") %>%
  str_flatten() %>%
  cat()
```

## Instructions

> Here are a number of personality traits that may or may not apply to you. Please write a number next to each statement to indicate the extent to which you agree or disagree with that statement. You should rate the extent to which the pair of traits applies to you, even if one characteristic applies more strongly than the other.

## Response options

-   Disagree strongly
-   Disagree moderately
-   Disagree a little
-   Neither agree nor disagree
-   Agree a little
-   Agree moderately
-   Agree strongly

## Download

```{r}
#| label: table
#| tbl-cap: TIPI items

items %>%
  select(where(~ sum(is.na(.x)) != length(.x))) %>%
  datatable(
    rownames = FALSE, 
    extensions = 'Buttons', 
    options = list(
      dom = 'Bip',
      buttons = list(
        'copy', 
        list(
          extend = 'collection',
          buttons = c('csv', 'excel'),
          text = 'Download'
        )
      )
    )
  )
```

## Reference

::: {#refs}
:::
