---
title: "New Ecological Paradigm Scale"
execute:
  echo: false
format: 
   html:
     df-print: kable
     toc: true
knitr:
  opts_chunk: 
    R.options:
      knitr.kable.NA: ''
nocite: |
  @dunlap2000
bibliography: references.bib
---

```{r}
#| label: setup
#| message: false
library(tidyverse)
library(DT)
items <- read_csv("NEP-items.csv")
```

The NEP is a 15-item measure of proenvironmental orientation. The revised NEP scale was designed to improve upon the original one in several respects:

1.  It taps into a wider range of facets of an ecological worldview
2.  It offers a balanced set of pro- and anti-NEP items
3.  It avoids outmoded terminology

## Items

```{r}
#| results: asis
items %>%
  select(item_nr, item) %>%
  unite(col = item, everything(), sep = ". ") %>%
  pull(item) %>%
  str_c("\n") %>%
  str_flatten() %>%
  cat()
```

## Instructions

> Listed below are statements about the relationship between humans and the environment. For each one, please indicate whether you STRONGLY AGREE, MILDLY AGREE, are UNSURE, MILDLY DISAGREE or STRONGLY DISAGREE with it.

## Response options

-   Strongly agree
-   Mildly agree
-   Unsure
-   Mildly disagree
-   Strongly disagree

## Download

```{r}
#| label: table
#| tbl-cap: NEP items

items %>%
  select(where(~ sum(is.na(.x)) != length(.x))) %>%
  datatable(
    rownames = FALSE, 
    extensions = 'Buttons', 
    options = list(
      dom = 'Bip',
      buttons = list(
        'copy', 
        list(
          extend = 'collection',
          buttons = c('csv', 'excel'),
          text = 'Download'
        )
      )
    )
  )
```

## Reference

::: {#refs}
:::