---
title: "The Optimism–Pessimism Short Scale–2"
execute:
  echo: false
format: 
   html:
     df-print: kable
     toc: true
knitr:
  opts_chunk: 
    R.options:
      knitr.kable.NA: ''
nocite: |
  @nießen2022
bibliography: references.bib
---

```{r}
#| label: setup
#| message: false
library(tidyverse)
library(DT)
items <- read_csv("SOP2-items.csv")
```

The Optimism--Pessimism Short Scale--2 (SOP2) is a 2-item measure of optimism---the tendency to look to the future with confidence and mostly expect good things to happen, rather than to look at the future full of doubt and mostly expect bad things to happen.

## Items

```{r}
#| results: asis
items %>%
  select(item_nr, item) %>%
  unite(col = item, everything(), sep = ". ") %>%
  pull(item) %>%
  str_c("\n") %>%
  str_flatten() %>%
  cat()
```

## Response options

Each question is rated on a 7-point scale, but only the first and last option is labelled.

For the optimism item:

-   not at all optimistic
-   very optimistic

For the pessimism item:

-   not at all pessimistic
-   very pessimistic

## Download

```{r}
#| label: table
#| tbl-cap: Speciesism Scale items

items %>%
  select(where(~ sum(is.na(.x)) != length(.x))) %>%
  datatable(
    rownames = FALSE, 
    extensions = 'Buttons', 
    options = list(
      dom = 'Bip',
      buttons = list(
        'copy', 
        list(
          extend = 'collection',
          buttons = c('csv', 'excel'),
          text = 'Download'
        )
      )
    )
  )
```

## Reference

::: {#refs}
:::